---
title: "Brewster_Analysis"
author: "Sean Collins"
date: "December 9, 2015"
output: html_document
---
Analysis for Brewster's physiology of exercise lab report 

Using the NFL Combine Data (Freely available online): http://nflcombineresults.com/nflcombinedata.php

Decided to use "all40yard" - not just the NFL combined timed sprints

Goal: test the relationship between the 3 cone test, 40 yard dash times and power produced from vertical jump

Will stick with the Sayers peak power from the vertical jump


### Analysis approach
1. Load, select and clean data relevant for this analysis
2. Descriptive statistics of relevant data
3. Univariate distributions of relevant data
4. Bivariate distributions with scatterplots
5. Linear regression

For the sake of presentation 4 & 5 will be combined so that output releated to the linear regression model will follow the scatterplot of the same variables.


## Setting working directory and loading required R packages:
```{r}
setwd("~/physexlab/NFLcombineData")
library(dplyr) #For data manipulation
library(ggplot2) #For plotting
library(pastecs) #For descriptive statistics
library(psych) #For descriptive statistics
library(gtools)

```

## Load, select and clean data relevant for this analysis

```{r}
nfldata<-read.csv("data_cleaned.csv")
nfldata<-mutate(nfldata, Weight_kg = Weight_lbs/2.2)
nfldata <- mutate(nfldata, Height_m = Height_in*0.0254)
nfldata <- mutate(nfldata, VertJump_m = VertLeap_in*0.0254)
nfldata <- mutate(nfldata, LewisPower = sqrt(4.9)*Weight_kg*9.81*sqrt(VertJump_m))
nfldata <- mutate(nfldata, SayersPower = (60.7*(VertJump_m*100))+(45.3*Weight_kg-2055))

#Select variables of interest 
nfldata <- select(nfldata, Year, POS, Weight_lbs, X3Cone, all40yard, SayersPower)

#Keep only complete cases to avoid missing values
nfldata <- nfldata[complete.cases(nfldata), ]
```

## Summary of data after the above process
```{r}
str(nfldata)
summary(nfldata)
```
3590 complete cases with the variables of interest, all positions (1999 - 2015)

## Descriptive statistics of relevant data
```{r}
options(scipen=100) #supresses scientific notation of output
options(digits=2) #suggests a limit of 2 significant digits - but it is only a suggestion
stat.desc(nfldata, basic=F) #runs basic stats
```

## Univariate distributions of relevant data
```{r}
qplot(X3Cone, data=nfldata, geom="histogram")
qplot(all40yard, data=nfldata, geom="histogram")
qplot(SayersPower, data=nfldata, geom="histogram")
qplot(Weight_lbs, data=nfldata, geom="histogram")
```

All well normally distributed - except for weight in pounds. There seem to be four distinct groups of weight in the draft. I will explore this more after doing the primary analysis.

## Bivariate distributions with scatterplots and Linear regression


### X3Cone and 40 yard dash
```{r}
qplot(y=all40yard, x=X3Cone, data=nfldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
X3Cone_SprintFit <- lm(all40yard ~ X3Cone, data=nfldata) #fit model
summary(X3Cone_SprintFit) # show results 
```

### Sayers Power (peak power from vertical jump)
```{r}
qplot(y=all40yard, x=SayersPower, data=nfldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
SayersPower_SprintFit <- lm(all40yard ~ SayersPower, data=nfldata) #fit model
summary(SayersPower_SprintFit) # show results 
```

### Sayers Power (peak power from vertical jump) and 40 yard dash
```{r}
qplot(y=all40yard, x=SayersPower, data=nfldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
SayersPower_SprintFit <- lm(all40yard ~ SayersPower, data=nfldata) #fit model
summary(SayersPower_SprintFit) # show results 
```

### Sayers Power (peak power from vertical jump) and X3Cone
```{r}
qplot(y=X3Cone, x=SayersPower, data=nfldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
SayersPower_X3ConeFit <- lm(X3Cone ~ SayersPower, data=nfldata) #fit model
summary(SayersPower_X3ConeFit) # show results 
```

### Multivariable regression Does combining X3Cone and sayers power predict 40 yard dash?
```{r}
qplot(y=all40yard, x=SayersPower, data=nfldata, geom=c("point", "smooth"),method="lm",size = X3Cone)
```

```{r}
SayersX3Cone_SprintFit <- lm(all40yard ~ SayersPower + X3Cone, data=nfldata) #fit model
summary(SayersX3Cone_SprintFit) # show results 
```
With both sayers and X3Cone in the regression model they are both significant preditors of sprint; and the overall R^2 is pretty good, but not much better than just the cone test alone.


### What about weight?
As pointed out earlier there are two distinct groups of WR weights based on the histogram:
```{r}
qplot(Weight_lbs, data=nfldata, geom="histogram")
```

Does weight contribute to our model of 40 yard sprint time?
```{r}
qplot(y=all40yard, x=SayersPower, data=nfldata, geom=c("point", "smooth"),method="lm",size = X3Cone, color = Weight_lbs)
```

```{r}
SayersX3ConeWeight_SprintFit <- lm(all40yard ~ SayersPower + X3Cone + Weight_lbs, data=nfldata) #fit model
summary(SayersX3ConeWeight_SprintFit) # show results 
```

Knowing body weight increases the overall R^2 significantly; and weight is a significant predictor in the model in the direction you would expect, the more weight the greater the time (greater time means slower)

#### Can this be made more clear? 
Another way to try this is to create a factor (category) variable based on the weight distribution. The code below breaks weight into 2 categories based on the distribution.
```{r}
nfldata$wt_quantiles<-quantcut(nfldata$Weight_lbs, q=seq(0,1,by=0.25), na.rm=TRUE)
```

We can see that this process does a nice job of partitioning the histogram from earlier into 4 discrete categories of weight. 
```{r}
qplot(Weight_lbs, data=nfldata, geom="histogram",color=nfldata$wt_quantiles)
```

Now the scatter plot with the weight grouping:
```{r}
qplot(y=all40yard, x=SayersPower, data=nfldata, geom=c("point", "smooth"),method="lm",size = X3Cone, color = nfldata$wt_quantiles)
```

```{r}
SayersX3ConeWeight2_SprintFit <- lm(all40yard ~ SayersPower + X3Cone + wt_quantiles, data=nfldata) #fit model
summary(SayersX3ConeWeight2_SprintFit) # show results 
```
While this factor (as opposed to continuous) approach is better visually with the scatter plot; it is not better mathematically. The factor variable for weight accounts for less of the variability in the data (lower R^2), which makes sense. 


#### To get a sense of positional differences in the variable above

Here are boxplots of each variable and how it varies by position - the black line in the middle of the plot is the overall mean; the upper and lower red lines are the +/- 1 Standard Deviation values.
##### 40 yard dash

```{r}
ggplot(nfldata, aes(factor(POS), all40yard)) + geom_boxplot() + geom_hline(aes(yintercept=mean(all40yard))) + geom_hline(aes(yintercept=(mean(all40yard)-sd(all40yard))), colour="#BB0000") + geom_hline(aes(yintercept=(mean(all40yard)+sd(all40yard)), colour="#BB0000"))
```

##### Sayers Power

```{r}
ggplot(nfldata, aes(factor(POS), SayersPower)) + geom_boxplot() + geom_hline(aes(yintercept=mean(SayersPower))) + geom_hline(aes(yintercept=(mean(SayersPower)-sd(SayersPower))), colour="#BB0000") + geom_hline(aes(yintercept=(mean(SayersPower)+sd(SayersPower)), colour="#BB0000"))
```

##### Cone test

```{r}
ggplot(nfldata, aes(factor(POS), X3Cone)) + geom_boxplot() + geom_hline(aes(yintercept=mean(X3Cone))) + geom_hline(aes(yintercept=(mean(X3Cone)-sd(X3Cone))), colour="#BB0000") + geom_hline(aes(yintercept=(mean(X3Cone)+sd(X3Cone)), colour="#BB0000"))
```

##### Weight (lbs)

```{r}
ggplot(nfldata, aes(factor(POS), Weight_lbs)) + geom_boxplot() + geom_hline(aes(yintercept=mean(Weight_lbs))) + geom_hline(aes(yintercept=(mean(Weight_lbs)-sd(Weight_lbs))), colour="#BB0000") + geom_hline(aes(yintercept=(mean(Weight_lbs)+sd(Weight_lbs)), colour="#BB0000"))
```

### Final thoughts -

Seems to be some interesting findings to discuss - let me know if you have any questions, or based on the insights of this analysis whether you have any other questions that I might have time to do. 