---
title: "VerticalJump-Inches-v-Power(NFL CombineData)"
author: "Sean Collins"
date: "December 4, 2015"
output: html_document
---
Decided to put together this analysis to demonstrate a few concepts. 
First, that it is important to analyze vertical jump based on power, not inches or centimeters. This is true due to the wide variation in power associated with the subjects body mass.
Second, it is a nice demonstration of what we mean by "explaining variance" in a regression when looking at scatter plots to see bivariate distributions.

For this demonstration we are going to look at the NFL Combine Data (Freely available online): http://nflcombineresults.com/nflcombinedata.php

For the sake of anyone interested I have left the "echo" feature on "TRUE" so the r code used to do this analysis is available for anyone interested. If you are not interested you can simply ignore the grey code boxes.

Prior to any of this work I had to scrap the data from the web page (that is separte code from what is here to save space and time). If anyone is interested in that code just let me know. Here I start with a csv file created from the XML webdata.

## Setting working directory and loading required R packages:
```{r}
setwd("~/physexlab/NFLcombineData")
library(dplyr)
library(ggplot2)
library(gtools)

```

## Loading the data and calculating derived variables:

```{r}
nfldata<-read.csv("data_cleaned.csv")
nfldata<-mutate(nfldata, Weight_kg = Weight_lbs/2.2)
nfldata <- mutate(nfldata, Height_m = Height_in*0.0254)
nfldata <- mutate(nfldata, BMI = Weight_kg / (Height_m**2))
nfldata <- mutate(nfldata, VertJump_m = VertLeap_in*0.0254)
nfldata <- mutate(nfldata, LewisPower = sqrt(4.9)*Weight_kg*9.81*sqrt(VertJump_m))
nfldata <- mutate(nfldata, SayersPower = (60.7*(VertJump_m*100))+(45.3*Weight_kg-2055))

```

## Plots of Power (Y axis) and Vertical Leap in inches (X Axis)
Note for both of these scatter plots it is clear there is significant vertical variation (for any particular height jumped players vary significantly in power)

### Sayers Power is an estimate of the Peak Power achieved during the jump 
```{r}
qplot(VertLeap_in, SayersPower, data = nfldata)
```

### Lewis Power is an estimate of the Mean Power during the jump
```{r}
qplot(VertLeap_in, LewisPower, data = nfldata)
```

## Why the wide variation?
In this sample of NFL Combine participants (n = 5394 between 1999 - 2015) there is significant variability in body mass. Also note that the variation is multi modal (there are several peaks)

```{r, echo=FALSE}
qplot(Weight_kg, data=nfldata, geom="histogram")
```

## How does this variation map to the bivariate variation we see in the scatter plots?
To answer this question we can first color each point in the bivariate distribution by body mass. It is clear from these plots that weight has a major influence (not surprising given the equations, and well, physics). 


### Sayers Power is an estimate of the Peak Power achieved during the jump 
```{r}
qplot(VertLeap_in, SayersPower, data = nfldata, color = nfldata$Weight_kg)
```

### Lewis Power is an estimate of the Mean Power during the jump
```{r}
qplot(VertLeap_in, LewisPower, data = nfldata, color = nfldata$Weight_kg)
```

### Can this be made more clear? 
To make what is happening even more clear we can create a factor (category) variable based on the weight distribution. The code below breaks weight into 4 categories based on the distribution.
```{r}
nfldata$wt_quantiles<-quantcut(nfldata$Weight_kg, q=seq(0,1,by=0.25), na.rm=TRUE)
```

We can see that this process does a nice job of partitioning the histogram from earlier into 4 discrete categories of weight. 
```{r}
qplot(Weight_kg, data=nfldata, geom="histogram",color=nfldata$wt_quantiles)
```

Now we can repeat the scatterplots (bivariate distributions) with the point color by these discrete body mass groups:

### Sayers Power is an estimate of the Peak Power achieved during the jump 
```{r}
qplot(VertLeap_in, SayersPower, data = nfldata, color = nfldata$wt_quantiles)
```

### Lewis Power is an estimate of the Mean Power during the jump
```{r}
qplot(VertLeap_in, LewisPower, data = nfldata, color = nfldata$wt_quantiles)
```

It is clear from both of the above graphics that there is a large impact of body mass, and for th most part it simply partitions the data into four separate positive relationships.


Sayers:
```{r}
qplot(y=SayersPower, x=VertLeap_in, col=factor(wt_quantiles), data=nfldata) + stat_smooth(method=lm, formula=y~x)
```


Lewis:
```{r}
qplot(y=LewisPower, x=VertLeap_in, col=factor(wt_quantiles), data=nfldata) + stat_smooth(method=lm, formula=y~x)
```

### From scatter plots to regression:

#### Sayers regression first without controlling for weight; then controlling for weight
Not controlling for weight - R^2 = 0.005033
```{r}
fitSayers<-lm(formula = SayersPower ~ VertLeap_in, data = nfldata)
summary(fitSayers)
```

Controlling for weight - R^2 = 0.8977
```{r}
fitSayersWeight<-lm(formula = SayersPower ~ VertLeap_in, data = nfldata)
summary(fitSayersWeight)
```

#### Lewis regression first without controlling for weight; then controlling for weight

Not controlling for weight - R^2 = 0.1125
```{r}
fitLewis<-lm(formula = LewisPower ~ VertLeap_in, data = nfldata)
summary(fitLewis)
```

Controlling for weight - R^2 = 0.9072
```{r}
fitLewisWeight<-lm(formula = LewisPower ~ VertLeap_in + wt_quantiles, data = nfldata)
summary(fitLewisWeight)
```
