---
title: "Anaerobic Power - NHL draft combine data"
author: "Sean Collins"
date: "December 10, 2015"
output: html_document
---
## Anaerobic Power - NHL draft combine data

### Draft analysis 

#### Ideas that are being tested
1. Relationship between peak HR on Wingate and peak HR on VO2 max bike test
2. Relationship between Wingate HR 1 min post and peak HR on VO2 max bike test
3. Relationship between peak HR on Wingate and Wingate HR 1 min post
3. Relationship between peak Watts on Wingage and peak Watts on VO2 max bike test

```{r}
# Load necessary R packages
library(dplyr) #For data manipulation
library(reshape2) #For data manipulation related to plotting
library(ggplot2) #For plotting
library(pastecs) #For descriptive statistics

# Set working directory - mac file system
setwd("~/physExLab/NHLcombinedata")

# Load data
nhldata<-read.csv("data.csv")

#Create weight in kg variable
nhldata<-mutate(nhldata, Weight_kg = Weight_lb/2.2)

#Create aerobic test peak watts / kg variable
nhldata<-mutate(nhldata, VO2max_watts_kg= VO2max_watts/Weight_kg)

#Create variable for how much the HR dropped after the Wingate
nhldata<-mutate(nhldata, rHR_wingate = WingateHR_End - WingateHR_1minPost)

#Create dataframe with only the variables we are interested in
nhldata<-select(nhldata,Weight_kg, WingatePeak_Watts, WingateMean_Watts, PeakWingate_W_kg, VO2max_l_min, VO2max_ml_kg_min, VO2max_watts, VO2max_watts_kg,VO2max_PeakHR, WingatePercentDrop, WingateHR_End, WingateHR_1minPost, rHR_wingate, SayersPeak_NoPause_watts, LewisAvg_.NoPause_watts)

nhldata <- nhldata[complete.cases(nhldata), ]
```

#### Descriptive stats of variables of interest
Note: only subjects with complete cases are included

```{r}
options(scipen=100) #supresses scientific notation of output
options(digits=2) #limits output to 2 significant digits
stat.desc(nhldata, basic=F) #runs basic stats
```

#### Univariate distributions with histograms

```{r}
qplot(Weight_kg, data=nhldata, geom="histogram")
qplot(WingatePeak_Watts, data=nhldata, geom="histogram")
qplot(WingateMean_Watts, data=nhldata, geom="histogram")
qplot(PeakWingate_W_kg, data=nhldata, geom="histogram")
qplot(VO2max_l_min, data=nhldata, geom="histogram")
qplot(VO2max_ml_kg_min, data=nhldata, geom="histogram")
qplot(VO2max_watts, data=nhldata, geom="histogram")
qplot(VO2max_PeakHR, data=nhldata, geom="histogram")
qplot(WingatePercentDrop, data=nhldata, geom="histogram")
qplot(WingateHR_End, data=nhldata, geom="histogram")
qplot(WingateHR_1minPost, data=nhldata, geom="histogram")
qplot(rHR_wingate, data=nhldata, geom="histogram")
qplot(SayersPeak_NoPause_watts, data=nhldata, geom="histogram")
qplot(LewisAvg_.NoPause_watts, data=nhldata, geom="histogram")
``` 

#### Relationship between peak HR on Wingate and peak HR on VO2 max bike test
```{r}
qplot(y=WingateHR_End, x=VO2max_PeakHR, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
WingateHR_VO2HR_Fit <- lm(WingateHR_End ~ VO2max_PeakHR, data = nhldata)
summary(WingateHR_VO2HR_Fit)
```

#### Relationship between Wingate HR 1 min post and peak HR on VO2 max bike test
```{r}
qplot(y=WingateHR_1minPost, x=VO2max_PeakHR, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
WingateHRpost_VO2HR_Fit <- lm(WingateHR_1minPost ~ VO2max_PeakHR, data = nhldata)
summary(WingateHRpost_VO2HR_Fit)
```

#### Relationship between peak HR on Wingate and Wingate HR 1 min post
```{r}
qplot(y=WingateHR_1minPost, x=WingateHR_End, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
WingateHRpost_VO2HR_Fit <- lm(WingateHR_1minPost ~ WingateHR_End, data = nhldata)
summary(WingateHRpost_VO2HR_Fit)
```

#### Is there a better relationship between vertical jump power (peak and average) and the wingate power (peak and average) after subtracting what could be considered the aerobic component of power?

```{r}
qplot(y=WingatePeak_Watts, x=SayersPeak_NoPause_watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
qplot(y=WingateMean_Watts, x=LewisAvg_.NoPause_watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

Create variables
```{r}
nhldata<-mutate(nhldata, WingatePeakW_minusAerobicWatts= WingatePeak_Watts - VO2max_watts)
nhldata<-mutate(nhldata, WingateMeanW_minusAerobicWatts= WingateMean_Watts - VO2max_watts)
```

```{r}
qplot(y=WingatePeakW_minusAerobicWatts, x=SayersPeak_NoPause_watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

```{r}
qplot(y=WingateMeanW_minusAerobicWatts, x=LewisAvg_.NoPause_watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```
