---
title: "Almonte - NHL draft combine data"
author: "Sean Collins"
date: "December 8, 2015"
output: html_document
---
Data analysis performed for D'Amour's physiology of exercise lab report

"bench, pushups, and push power would be the best way to represent the "hitting strength" for the players. Id compare the two numbers for bench and pushups to factor in body weight too so its not looking at if heavier people are stronger as well""

Question: Do skating players (forward and defense) entering the NHL combine scouting tests have different testing results indices of "hitting strength"? 

Operationalize "hitting strength" as: bench, pushups, and push power; Consider both absolute values and normalize for body mass

Expected: Defense will have a higher "hitting strength"

Note: goalies are not included in the analysis

###Approach:
1. Load and clean data (take complete cases and variables of interest)
2. Descriptive statistics for each group
3. Boxplots to compare groups
4. t-tests to compare group differences (statistically test the null hypothesis that the groups are not different)

All analyses performed in R (version 3.2.0 (www.r-project.org)) using RStudio (www.rstudio.com)

####Variables
BenchPress_reps150
BenchPress_lbperlbBodyWeight
PushUps_max
PushUpsxBody.Weight
Push.Strength..lb.
PushStrength_lbperlbBodyWeight

Note: since data includes body mass normalized variables, this analysis will not calculate those variables; therefore the normalizations are based on "lbs" not "kgs" since that is how the NHL has done the processing

###Loading R packages, getting data ready
```{r}
# Load necessary R packages
library(dplyr) #For data manipulation
library(reshape2) #For data manipulation related to plotting
library(ggplot2) #For plotting
library(pastecs) #For descriptive statistics
library(psych) #For descriptive statistics

# Set working directory - mac file system
setwd("~/physExLab/NHLcombinedata")

# Load data
nhldata<-read.csv("data_position.csv")

#Create dataframe with only the variables we are interested in
nhldata<-select(nhldata,PosNumeric, Position, Weight_lb,BenchPress_reps150, BenchPress_lbperlbBodyWeight, PushUps_max, PushUpsxBody.Weight, Push.Strength..lb., PushStrength_lbperlbBodyWeight)
```

#### Loaded data with all forwards and defense (including subjects with missing values)
As you can see there are missing values in most variables.
```{r}
str(nhldata)
summary(nhldata)
```

But we are really only interested in cases that have complete data - so need to eliminate cases with missing data (note - if we had plans for a publication we would be much more cautious about doing this)

#### Cleaning loaded data to extract only "complete cases"
A complete case has all of the variables of interest

We have gone from 90 subjects to 82 subjects - so 8 subjects had incomplete data in the variables of interest.

```{r}
nhldata <- nhldata[complete.cases(nhldata), ]
str(nhldata)
summary(nhldata)
```

##### Descriptive summary (both groups):
```{r}
options(scipen=100) #supresses scientific notation of output
options(digits=2) #suggests a limit of 2 significant digits - but it is only a suggestion
stat.desc(nhldata, basic=F) #runs basic stats
```

##### Descriptive summary for forwards:
```{r}
forwards<-filter(nhldata,PosNumeric==1)
options(scipen=100) #supresses scientific notation of output
options(digits=2) #suggests a limit of 2 significant digits - but it is only a suggestion
stat.desc(forwards, basic=F) #runs basic stats
```

##### Descriptive summary for defenders:
```{r}
defense<-filter(nhldata,PosNumeric==2)
options(scipen=100) #supresses scientific notation of output
options(digits=2) #suggests a limit of 2 significant digits - but it is only a suggestion
stat.desc(defense, basic=F) #runs basic stats
```

###NOTE - LEFT OFF HERE

Decided to use Boxplots - The box plot (a.k.a. box and whisker diagram) is a standardized way of displaying the distribution of data based on the five number summary: minimum, first quartile, median, third quartile, and maximum. Outliers are plotted as individual points. They are easier to look at for comparing two factors of a variable (forward, defense) 

```{r}
ggplot(nhldata, aes(factor(Position), WingatePeak_Watts)) + geom_boxplot() 
```

```{r}
ggplot(nhldata, aes(factor(Position), PeakWingate_W_kg)) + geom_boxplot() 
```

```{r}
ggplot(nhldata, aes(factor(Position), WingatePercentDrop)) + geom_boxplot() 
```

```{r}
ggplot(nhldata, aes(factor(Position), VO2max_l_min)) + geom_boxplot()
```

```{r}
ggplot(nhldata, aes(factor(Position), VO2max_ml_kg_min)) + geom_boxplot()
```

```{r}
ggplot(nhldata, aes(factor(Position), VO2max_watts_kg)) + geom_boxplot()
```

```{r}
ggplot(nhldata, aes(factor(Position), LewisAvg_.Pause_watts)) + geom_boxplot()
```

```{r}
ggplot(nhldata, aes(factor(Position), LewisAvg_.Pause_watts_kg)) + geom_boxplot()
```

The fact that there is not much difference between the absolute and relative versions of these data is really due to the fact that there is not much of a difference between the groups in body mass:
```{r}
ggplot(nhldata, aes(factor(Position), Weight_kg)) + geom_boxplot()
```

Just for kicks - I also used a t-test to see whether there were differences between the positions (the graphics suggest not - but wanted to see if the null was accepted in the t-test (i.e. p > 0.05 if we accept an alpha of 0.05))

```{r}
t.test(nhldata$Weight_kg~nhldata$Position)
t.test(nhldata$WingatePeak_Watts~nhldata$Position)
t.test(nhldata$PeakWingate_W_kg~nhldata$Position)
t.test(nhldata$WingatePercentDrop~nhldata$Position)
t.test(nhldata$VO2max_l_min~nhldata$Position)
t.test(nhldata$VO2max_ml_kg_min~nhldata$Position)
t.test(nhldata$VO2max_watts~nhldata$Position)
t.test(nhldata$VO2max_watts_kg~nhldata$Position)
t.test(nhldata$LewisAvg_.Pause_watts~nhldata$Position)
t.test(nhldata$LewisAvg_.Pause_watts_kg~nhldata$Position)
```

Clearly there are no significant differences in this data set between these players. Makes for an interesting report though - particularly if there are papers out there showing a difference in older players. Raises the question of whether the players have not been playing long enough to be trained into their position where they partially adapt based on their role.