---
title: "Hunt - NHL draft combine data"
author: "Sean Collins"
date: "December 2, 2015"
output: html_document
---
### Data analysis performed for Hunt's physiology of exercise lab report

updated on december 4, 2015 - set echo to false in the R code to remove the code from the analysis; and added additional labels and explanation.

#### Question:

Looking at the NHL combine data from (2010), how does the participants musculoskeletal fitness, specifically their peak vertical leg power (sayers), relate to their VO2max, precisely ml/kg/min and peak power output, specifically watts/kg?

VO2max = Aerobic, stationary bike with weighted increments
Peak power output = Anaerobic, wingate

### Analysis

#### Preliminary Analysis

This first output is simply unavoidable to load packages - it is not part of the analysis but I cannot get it to go away.

```{r, echo=FALSE, results="hide"}
# Load necessary R packages
library(dplyr) #For data manipulation
library(reshape2) #For data manipulation related to plotting
library(ggplot2) #For plotting
library(pastecs) #For descriptive statistics
```

```{r, echo=FALSE}
# Set working directory - mac file system
setwd("~/physExLab/NHLcombinedata")

# Load data
nhldata<-read.csv("data.csv")

#Create weight in kg variable
nhldata<-mutate(nhldata, Weight_kg = Weight_lb/2.2)

#Create aerobic test peak watts / kg variable
nhldata<-mutate(nhldata, VO2max_watts_kg= VO2max_watts/Weight_kg)

#Create Sayer peak watts / kg
nhldata<-mutate(nhldata, SayersPeak_NoPause_watts_kg= SayersPeak_NoPause_watts/Weight_kg)

#Create dataframe with only the variables we are interested in
nhldata<-select(nhldata,Weight_kg, WingatePeak_Watts, PeakWingate_W_kg, VO2max_l_min, VO2max_ml_kg_min, VO2max_watts, VO2max_watts_kg, SayersPeak_NoPause_watts, SayersPeak_NoPause_watts_kg)
```

##### Summary of Data of interest in this analysis: it is clear there is missing data

```{r, echo=FALSE}
str(nhldata)
summary(nhldata)
```

Since we are only interested in cases that have complete data - so need to eliminate cases with missing data

As you can see we have gone from 100 subjects to 94 subjects - so 6 subjects had incomplete data in the variables of interest.

##### Summary of data with cases that have missing values removed

```{r, echo=FALSE}
nhldata <- nhldata[complete.cases(nhldata), ]
str(nhldata)
summary(nhldata)
```

##### Descriptive summary of the variables of interest:

```{r, echo=FALSE}
options(scipen=100) #supresses scientific notation of output
options(digits=2) #limits output to 2 significant digits
stat.desc(nhldata, basic=F) #runs basic stats
```

####  Univariate distributions

These are separated in groups of variables that have similiar Y axis scaling 

Boxplots - The box plot (a.k.a. box and whisker diagram) is a standardized way of displaying the distribution of data based on the five number summary: minimum, first quartile, median, third quartile, and maximum. Outliers are plotted as individual points.  

#### Data with the common Y axis of Watts:

```{r, echo=FALSE}
watts<-select(nhldata,SayersPeak_NoPause_watts, WingatePeak_Watts,VO2max_watts)
ggplot(melt(watts), aes(x = variable, y = value)) + geom_boxplot() 
```

##### Data with the common Y axis of Watts/kg:

```{r, echo=FALSE}
watts_kg<-select(nhldata,SayersPeak_NoPause_watts_kg, PeakWingate_W_kg, VO2max_watts_kg)
ggplot(melt(watts_kg), aes(x = variable, y = value)) + geom_boxplot() 
```

##### Data with the common Y axis of oxygen consumption (one for L/min and one for L/kg/min):

```{r, echo=FALSE}
vo2l<-select(nhldata,VO2max_l_min)
vo2lkg<-select(nhldata,VO2max_ml_kg_min)
ggplot(melt(vo2l), aes(x = variable, y = value)) + geom_boxplot() 
ggplot(melt(vo2lkg), aes(x = variable, y = value)) + geom_boxplot() 
```

#### Bivariate distributions (scatter plots to look at associations between two variables)

Next we can compare bivariate distributions (i.e. scatter plots) to look at associations. This first draft will only include the scatter plots without assumptions of the best approach to fitting the data with a function (i.e. linear model).

##### VO2max in L/min and in L/kg/min:

```{r, echo=FALSE}
qplot(VO2max_l_min,VO2max_ml_kg_min,data=nhldata,geom=c("point"))
```

To help visualize what is happenning (why there is variability) we can weight the points based on the players weight - so that larger points indicate a heavier player. As you can see - heavier players may have larger VO2 max in L/min, but they have lower VO2 max in ml/kg/min - which makes sense.

```{r, echo=FALSE}
qplot(VO2max_l_min,VO2max_ml_kg_min,data=nhldata,geom=c("point"),size = Weight_kg)
```

##### VO2 L/min and VO2 watts

Watts on the X axis because technically the power being produced is a determinant of the oxygen need (and therefore consumed). One thing to note is that VO2 max watts seems to be a factor variable (not continuous), this is because of the mode of testing the players. They use a Monark bike and each stage of the max test is a pre determined power. So the max watts achieved is based on the state. Therefore, variation in actual watts due to variations in pedal RPMs from the RPMs required in the test would not be captured.

```{r, echo=FALSE}
qplot(VO2max_watts, VO2max_l_min,data=nhldata,geom=c("point"))
```

###### VO2 L/min and VO2 watts (weighted for body mass)

Here we look at the same plot as above but with the points weighted based on player weight. There is a much less clear pattern than the plot of VO2 L/min and VO2 ml/kg/min when weighting based on weight.
```{r, echo=FALSE}
qplot(VO2max_watts, VO2max_l_min,data=nhldata,geom=c("point"),size = Weight_kg, color = WingatePeak_Watts)
```

```{r, echo=FALSE}
qplot(VO2max_watts, VO2max_l_min,data=nhldata,geom=c("point"),size = Weight_kg, color = SayersPeak_NoPause_watts)
```

##### Wingate power - comparision between Watts and Watts/Kg:

```{r, echo=FALSE}
qplot(WingatePeak_Watts, PeakWingate_W_kg,data=nhldata,geom=c("point"))
```

##### Wingate power comparision between Watts and Watts/Kg

When weighting the points based on weight, just like in VO2, it is clear that those that have more mass have greater Watts, but not necessarily greater Watts/Kg

```{r, echo=FALSE}
qplot(WingatePeak_Watts, PeakWingate_W_kg,data=nhldata,geom=c("point"),size = Weight_kg)
```


##### Vertical jump average power watts to watts/kg

Expect the same as above - but a useful exercise in getting familiar with the data. 

###### first without weighting based on body mass

```{r, echo=FALSE}
qplot(SayersPeak_NoPause_watts, SayersPeak_NoPause_watts_kg,data=nhldata,geom=c("point"))
```

###### next with weighting based on body mass

```{r, echo=FALSE}
qplot(SayersPeak_NoPause_watts, SayersPeak_NoPause_watts_kg,data=nhldata,geom=c("point"),size = Weight_kg)
```

#### Primary analysis questions:

1. Compare Wingate peak watts to VO2 max watts 

```{r, echo=FALSE}
qplot(WingatePeak_Watts, VO2max_watts,data=nhldata,geom=c("point"))
```

2. Compare VO2 max ml kg min to peak wingate watt kg

```{r, echo=FALSE}
qplot(PeakWingate_W_kg, VO2max_ml_kg_min,data=nhldata,geom=c("point"))
```

3. Compare Sayers peak no pause watts to the peak watts and to VO2 max watts to see if there is any relationships and if jump capacity has an impact 

Expanded and improved upon
First with straight forward scatter plots for each comparision
Sayers peak no pause watts to the peak wingate watts - with visualization of the impact of weight
```{r, echo=FALSE}
qplot(SayersPeak_NoPause_watts, WingatePeak_Watts,data=nhldata,geom=c("point", "smooth"),method="lm",size = Weight_kg)
```

### Linear model
```{r}
fit <- lm(WingatePeak_Watts ~ SayersPeak_NoPause_watts, data=nhldata) #fit model
summary(fit) # show results 
```


Sayers peak no pause watts to the peak aerobic watts -with visualization of the impact of weight
```{r, echo=FALSE}
qplot(SayersPeak_NoPause_watts, VO2max_watts,data=nhldata,geom=c("point"),size = Weight_kg)
```

4. Next we want to see if there is a pattern in the distribution of vertical jump power across the bivariate distribution of wingate and aerobic power
```{r, echo=FALSE}
qplot(WingatePeak_Watts, VO2max_watts,data=nhldata,geom=c("point"),size = SayersPeak_NoPause_watts)
```

5. Finally we want to see if there is a pattern in the distribution of vertical jump power and body mass across the bivariate distribution of wingate and aerobic power - this may be too much in one graphic but it is a start:
```{r, echo=FALSE}
qplot(WingatePeak_Watts, VO2max_watts,data=nhldata,geom=c("point"),size = SayersPeak_NoPause_watts, color = Weight_kg)
```

There could be more to come. Let's start with the above and see where it takes the report. It may be interesting to put some of these into linear models and see what we get - but I want to do so cautiously, not just throw a model together to see what comes out of it. Once models are built, then the fit lines can be added to any / all of the graphics above.


