---
title: "Turcotte Analysis - NHL draft combine data"
author: "Sean Collins"
date: "December 9, 2015"
output: html_document
---
Data analysis performed for Turcott's physiology of exercise lab report

```{r}
# Load necessary R packages
library(dplyr) #For data manipulation
library(reshape2) #For data manipulation related to plotting
library(ggplot2) #For plotting
library(pastecs) #For descriptive statistics

# Set working directory - mac file system
setwd("~/physExLab/NHLcombinedata")

# Load data
nhldata<-read.csv("data.csv")

#Create weight in kg variable
nhldata<-mutate(nhldata, Weight_kg = Weight_lb/2.2)

#Create aerobic test peak watts / kg variable
nhldata<-mutate(nhldata, VO2max_watts_kg= VO2max_watts/Weight_kg)

#Create variable for how much the HR dropped after the Wingate
nhldata<-mutate(nhldata, rHR_wingate = WingateHR_End - WingateHR_1minPost)

#Create dataframe with only the variables we are interested in
nhldata<-select(nhldata,Weight_kg, WingatePeak_Watts, WingateMean_Watts, PeakWingate_W_kg, VO2max_l_min, VO2max_ml_kg_min, VO2max_watts, VO2max_watts_kg, WingatePercentDrop, WingateHR_End, WingateHR_1minPost, rHR_wingate)

nhldata <- nhldata[complete.cases(nhldata), ]
str(nhldata)
summary(nhldata)
```

As you can see we have gone from 100 subjects to 94 subjects - so 6 subjects had incomplete data in the variables of interest.

Descriptive summary of the variables of interest:
```{r}
options(scipen=100) #supresses scientific notation of output
options(digits=2) #limits output to 2 significant digits
stat.desc(nhldata, basic=F) #runs basic stats
```

#### Univariate distributions with histograms

```{r}
qplot(VO2max_ml_kg_min, data=nhldata, geom="histogram")
qplot(WingatePercentDrop, data=nhldata, geom="histogram")
qplot(WingateHR_1minPost, data=nhldata, geom="histogram")
qplot(WingateHR_End, data=nhldata, geom="histogram")
qplot(rHR_wingate, data=nhldata, geom="histogram")
``` 

#### Does a higher VO2 max result in a greater drop in heart rate after the Wingate?
```{r}
qplot(y=rHR_wingate, x=VO2max_ml_kg_min, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```
Looks like there is no relationship.

#### Does a higher VO2 max result in lower 1 minute post heart rate after the Wingate?
```{r}
qplot(y=WingateHR_1minPost, x=VO2max_ml_kg_min, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```
Looks like there is no relationship.

#### How does post wingate HR relate to % drop in power during the wingate? 
```{r}
qplot(y=WingateHR_End, x=WingatePercentDrop, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```
Looks like there is not much of a relationship. But does their peak watts influence the result?


#### But does their peak watts influence the result?
This plot is a repeat of the above with Wingate peak watts influencing the size of the point. From the points it looks like people with a higher peak watt have a greater drop off.
```{r}
qplot(WingatePercentDrop, WingateHR_End,data=nhldata,geom=c("point", "smooth"),method="lm",size = WingatePeak_Watts)
```

#### Do players with a higher peak watts on Wingate have a higher drop off during the test?
```{r}
qplot(y=WingatePercentDrop, x=WingatePeak_Watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```
Seems to be the case. 
To me this suggest that % drop is not due to poor aerobic capacity; but due to such high power output that players cannot sustain it. 

#### How does wingate average power fit in to the above relationship?
```{r}
qplot(WingatePeak_Watts, WingatePercentDrop,data=nhldata,geom=c("point", "smooth"),method="lm",size = WingateMean_Watts)
```

It looks like mean watts may be related to peak watts (makes sense - a higher peak leads to a higher mean), but that it is not related to percent drop as the scatter of point size along the x axis is fairly consistent.

Here is a plot of mean watts and percent drop to confirm those suspicions:
```{r}
qplot(y=WingatePercentDrop, x=WingateMean_Watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```


And here we see the relationship between wingate mean watts and peak watts - relationships don't really get much better than this one....
```{r}
qplot(y=WingateMean_Watts, x=WingatePeak_Watts, data=nhldata) + stat_smooth(method=lm, formula=y~x)
```

#### Thoughts - 
Peak watts and mean watts are highly related; and peak watts and percent drop are also related. It makes me think that % drop which is supposed to be the fatigue index is really driven by having high peak power. In other words, it's not as much that someone cannot sustain their power; it is that people with really high power cannot sustain it - possibly because it is, but it's nature, unsustainable.

The fact that vo2 max does not relate to recovery heart rate is a bit perplexing to me....there is just a lot of Vo2 max variability and not much HR recovery in one minute following the Wingate - so there is little variability in the Wingate HR 1 minute post.
